\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-16 paper.tex
save

In this work, we propose to perform variational optimization
on ${\cal L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability
of $g$. More specifically, we consider a proposal distribution $q_\psi(\theta)$
over the parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align}
respectively over $\phi$ and $\psi$.
When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
obtained by sampling mini-batches of true and generated data, as ordinarily
done in stochastic gradient descent. When updating $g$, gradient estimates
of $\nabla_\psi U_g$ can be derived with forward simulations, as described in Eqn.~\ref{eqn:vo-grad}.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
\end{equation}
which we can approximate by sampling mini-batches of generated data
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
\end{equation}
for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_z$.

\glnote{
While the above gradient estimate $\tilde{\nabla}_\psi U_g$ can readily be plugged
into the adversarial training procedure, it may also exhibit a sampling
variance that is larger than necessary, hence making the optimization unstable.
However, one can exploit the fact that $-d(g(\mathbf{z}; \theta); \phi)$
is a composition where $-d(\tilde{\mathbf{x}})$ is
known and differentiable with respect to $\tilde{\mathbf{x}}$.
In practice, using
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
\end{equation}
works much better ite seems, but I am not sure to precisely understand why, nor whether this is correct...
It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q_\psi$, whose collection
of generated samples is optimized to approach the real data distribution $p_r$.
In particular, this formulation does not necessarily guarantee that the
proposal distribution $q_\psi$ will place its mass arbitrarily tight
around the parameters of interest $\theta^*$. \glnote{Describe regularization
penalty on the parameters of the proposal to enforce that.}

>>>>

In this work, we propose to perform variational optimization
on ${\cal L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability
of $g$. More specifically, we consider a proposal distribution $q_\psi(\theta)$
over the parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align}
respectively over $\phi$ and $\psi$.
When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
obtained by evaluating the exact and known gradient of $U_d$ over
mini-batches of true and generated data, as ordinarily
done in stochastic gradient descent. When updating $g$, gradient estimates
of $\nabla_\psi U_g$ can be derived with forward simulations, as described in Eqn.~\ref{eqn:vo-grad}.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
\end{equation}
which we can approximate with mini-batches of generated data
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
\end{equation}
for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_z$.

\glnote{
While the above gradient estimate $\tilde{\nabla}_\psi U_g$ can readily be plugged
into the adversarial training procedure, it may also exhibit a sampling
variance that is larger than necessary, hence making the optimization unstable.
However, one can exploit the fact that $-d(g(\mathbf{z}; \theta); \phi)$
is a composition where $-d(\tilde{\mathbf{x}})$ is
known and differentiable with respect to $\tilde{\mathbf{x}}$.
In practice, using
\begin{equation}
    \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
\end{equation}
works much better it seems, but I am not sure to precisely understand why, nor whether this is correct...
It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q_\psi$, whose joint collection
of generated samples is optimized to approach the real data distribution $p_r$.
In particular, this formulation does not necessarily guarantee that the
proposal distribution $q_\psi$ will place its mass arbitrarily tight
around the parameters of interest $\theta^*$. \glnote{Describe regularization
penalty on the parameters of the proposal to enforce that.}
